Missing data: Current practice in football research and recommendations for improvement

### 2020-09-04 paper/paper.Rmd
change wording in the intro.

2: University of New South Wales, Department of Statistics, School of
Mathematics and Statistics, Sydney, Australia

3: Monash University, Department of Econometrics and Business Statistics,
Melbourne, Australia

4: Australian Centre of Excellence for Mathematical and Statistical
Frontiers (ACEMS), Melbourne, Australia

@@

@@

Nicholas J Tierney

Monash University, Department of Econometrics and Business Statistics,
Melbourne, Australia

@@

Missing data are often unavoidable. The reason values go missing, along
with decisions made of how missing data are handled (deleted or
imputed), can have a profound effect on the validity, usability and
accuracy of a study's results. In this article, we aimed to: estimate
the proportion of studies in football research that included a missing
data statement, highlight several improvable practices in relation to
missing data, and provide recommendations for exploring, visualising and
reporting missingness. Football related articles, published in 2019 were
studied. A survey of 137 articles, sampled at random, was conducted to
determine whether a missing data statement was included. As expected, the
proportion of studies in football research that included a missing data
statement was low, at only X% (95% confidence interval; X% to X%); suggesting that 
missingness is seldom considered by researchers. We suspect that this result 
is consistent with the wider sports science literature. We recommend 
researchers describe the number and percentage of missing
values, including when there are no missing values. Exploratory analysis
should be conducted to explore missing values, and visualisations
describing missingness overall should be provided in the paper, or at
least supplementary materials. Missing values should almost always be
imputed, and imputation methods should be explored to ensure they are
appropriately representative.

@@

Missing data are values that should have been observed, but were not. Since the
values are unobserved, this can mean the population of interest might not be 
properly sampled, inducing bias into study results. Left undetected or poorly handled, missing data can undermine the validity and accuracy of research results
[@Sainani2015; @Nakagawa2008; @Sterne2009]. Data can go
missing many ways. For example, accidentally skipping a survey question,
equipment failure, or intentionally not recording values. Before
analysis, missing data must be handled, with values typically deleted or
imputed. There is no universal approach to handling missing data. 
Contextual factors--such as, the study design and objective, and
pattern of missingness--determine how missings should be handled, on a
case-by-case basis [@Sainani2015].

For clarity, we describe some examples of missingness. An Australian football data collector was taking lunch during the third quarter, so no data was recorded for this time period. Compare this to an AFL game being cancelled due to COVID19. There is no missing data because there is no data to be observed. If a study screens out individuals who do not meet inclusion criteria, this is not a missing value, because there is no intent for these values to be observed or measured.

>>>>

2: University of New South Wales, Department of Statistics, School of
Mathematics and Statistics, Sydney, Australia.

3: Monash University, Department of Econometrics and Business Statistics,
Melbourne, Australia.

4: Australian Centre of Excellence for Mathematical and Statistical
Frontiers (ACEMS), Melbourne, Australia.

@@

The authors declare no conflicts of interest.

@@

Nicholas J Tierney
Monash University, Department of Econometrics and Business Statistics,
Melbourne, Australia.
Address:
Email:

@@

Missing data are often unavoidable. The reason values go missing, along
with decisions made of how missing data are handled (deleted or
imputed), can have a profound effect on the validity, accuracy and usability 
of a study's results. In this article, we aimed to: estimate
the proportion of studies in football research that included a missing
data statement, highlight several practices to avoid in relation to
missing data, and provide recommendations for exploring, visualising and
reporting missingness. Football related articles, published in 2019 were
studied. A survey of 137 articles, sampled at random, was conducted to
determine whether a missing data statement was included. As expected, the
proportion of studies in football research that included a missing data
statement was low, at only X% (95% confidence interval; X% to X%); suggesting that 
missingness is seldom considered by researchers. It is suspected that this result 
is consistent with practice in the wider sports science literature. We recommend 
researchers describe the number and percentage of missing
values, including when there are no missing values. Exploratory analysis
should be conducted to explore missing values, and visualisations
describing missingness overall should be provided in the paper, or at
least supplementary materials. Missing values should almost always be
imputed, and imputation methods should be explored to ensure they are
appropriately representative.

@@

Missing data are values that should have been observed, but were not. Since the
values are unobserved, this can mean the population of interest might not be 
properly sampled, inducing bias into a study's results. Left undetected or poorly handled, 
missing data can undermine the validity and accuracy of research results
[@Sainani2015; @Nakagawa2008; @Sterne2009]. Data can go
missing many ways. For example, accidentally skipping a survey question,
equipment failure, or intentionally not recording values. Before
analysis, missing data must be handled, with values typically deleted or
imputed. There is no universal approach to handling missing data. 
Contextual factors--such as, the study design and objective, and
pattern of missingness--determine how missings should be handled, on a
case-by-case basis [@Sainani2015].

For clarity, we describe an example of missingness. An Australian Rules Football data collector 
was taking lunch during the third quarter, so no data was recorded for this time period. This
would be considered missing data. Compare this to the game being cancelled due to a pandemic. 
There is no missing data because there is no data to be observed. Similiarly, if a study screens out individuals who 
do not meet inclusion criteria, this is not a missing value, because there is no intent for these 
values to be observed or measured.
